Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-09 paper/paper.tex
Algebras and Coalgebras

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\catafree}[1]{\Lbanana #1 \Rbanana}
\newcommand{\anacofree}[1]{\Llens #1 \Rlens}
\newcommand{\hylo}[2]{\cata{#1 \to #2}}
\newcommand{\cohylo}[2]{\ana{#1 \to #2}}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\head}{\operator{head}}
\newcommand{\inl}{\operator{inl}}
\newcommand{\inr}{\operator{inr}}
\newcommand{\tail}{\operator{tail}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}
\newcommand{\fmap}[1]{\text{fmap}\;#1}

@@

\newcommand{\List}[1]{\operator{List}(#1)}
\newcommand{\nil}{\operator{nil}}
\newcommand{\cons}{\operator{cons}}
\newcommand{\foldr}{\operator{foldr}}

@@

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16, HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third, Morihata09:Third,
farmsCastro}).  These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

@@

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the work
in mechanising recursion schemes focuses on a narrow subset of the known
recursion schemes due to termination issues, or do not focus on proving
algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,
MurataE19, larchey2022braga}.  Indeed, most of the mechanisations of generic
recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
on the \emph{unfolds}. In general, Hinze et al.~\cite{HinzeWG15} has proven that
all known  recursion schemes are instances of a  divide-and-conquer algorithm
which goes under the name of \emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To
date, no work has mechanised hylomorphisms in the Coq proof assistant, together
with their algebraic laws.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration is what allows the execution of code that has been verified in
Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
al.'s approach, extraction (1) does not preserve the recursive structure of
common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
the generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

@@

% \section{Background on Recursion Schemes}
% Recursion schemes provide a systematic way to work with data types, such as
% trees or lists, by abstracting away the recursion into higher-order functions.
% Canonical recursion schemes include \emph{folds} (catamorphisms) which consume
% data and \emph{unfolds} (anamorphisms) which produce data. These are the unique
% solutions (when they exist) to the following two equations respectively
% \[
%   h  = a \comp F h \comp \InOp \qquad k = \OutOp \comp F k \comp c
% \]
% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on
% the greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When
% the unique solutions to the above equations exist we denote them as $\cata{a}$
% and $\ana{c}$ respectively.

% \subsection{Hylomorphisms}
% Catamorphisms (dually anamorphisms) do not capture every recursion
% scheme we can think of.  As an example, consider an example of a classic
% \emph{divide-and-conquer} program such as quicksort algorithm:\footnote[1]{This
% program is written in Coq notation to be consistent, but it is not a program
% that Coq can accept.}

% \mpav{I need this to be written in Coq notation}
% \begin{coqcode}
% qs :: [a] -> [a]
% qs [] = []
% qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
% \end{coqcode}
% This program is formed of essentially two parts. The first one is what is
% commonly thought as the \emph{partition} function destructuring the input list
% in three parts, the pivot and the other two sublists, and the second part
% recursing over the sublists and composing the results back together.

% Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
% the obvious functorial action then \coq{partition} can be viewed as a
% \coq{TreeF}-coalgebra on lists over the type $A$
% \begin{coqcode}
% c : [A] -> TreeF [A]
% c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
% \end{coqcode}

% While we can define an \coq{TreeF}-algebra on lists
% over $A$ as well, that is a map
% \begin{coqcode}
% a : TreeF [A] -> [A]
% a (left, x, right) = left ++ [x] ++ right
% \end{coqcode}

% It should be easy to see that the program \coq{qs} can now be defined as the
% unique program that runs the coalgebra $c$ recurses over the sublists produced
% and then composes the results together. The structure exemplified by this
% program is a typical divide-and-conquer algorithm which can be implemented using
% an \emph{hylomorphism}. An hylomorphism is the unique solution $x : C \to A$ to
% the hylo equation
% \begin{equation}
%   \label{eq:hylo}
%   x =  a \comp Fx \comp c
% \end{equation}
% for a $F$-algebra on $A$, an $F$-coalgebra on $C$ and a functor $F$.  If for
% every $F$-algebra there exists a unique solution to (\ref{eq:hylo}) then we say
% the $F$-coalgebra $c$ is \emph{recursive}. Dually if for every $F$-coalgebra $c$
% there exists a unique solution to (\ref{eq:hylo}) we say the $F$-algebra $a$ is
% \emph{corecursive}. We denote the unique solution to the hylo equation
% (\ref{eq:hylo}) as $\hylo{a}{c}$.

% Moreover, it shoud be easy to see that catamorphisms and anamorphism arise as a
% special case of hylomorphisms
% \begin{equation}
%   \label{eq:cata}
%   \cata{a} = \hylo{\InOp}{a} \qquad \ana{c} = \hylo{c}{\OutOp}
% \end{equation}
% since $\InOp$ is indeed a recursive coalgebra and $\OutOp$ is a corecursive
% algebra.

% \subsection{Implementing Recursion Schemes}
% Hylomorphisms can implement all structured recursion schemes~\cite{HinzeWG15,
% HinzeW16}.  Here we show the result by means of a very basic example: mutual
% recursion. We refer the reader to Hinze et al.~\cite{HinzeWG15} for the
% theoretical result and to Yang and Wu~\cite{YangW22} for more examples of
% recursion schemes.

% % More specifically, in the context of recursive data types, such as in Haskell,
% % where induction and coinduction is conflated, an hylomorphism can be implemented
% % in terms of a catarmorphism (folding) and an anamorphism (unfolding)
% % \[
% %   \hylo{a}{c} = \cata{a} \comp \ana{c}
% % \]
% % They first generate a data structure using an anamorphism and then immediately
% % fold or reduce it using a catamorphism.

% % The hylomorphism pattern is particularly useful for defining recursive
% % algorithms in a functional programming context, where recursion is a fundamental
% % tool for solving problems. By separating the generation of data structures from
% % their consumption, hylomorphisms enable clearer and more modular code, making it
% % easier to reason about and maintain recursive algorithms.

% % However, in a type theory inductive and coinductive data types are separate.

% % In this section we will introduce the notion of hylomorphism. By way of examples
% % we will also give high level intution as to way all recursion schemes can be
% % defined in terms of a basic

% Say that we want to prove that the two mutually recursive functions below are
% well-defined

% \begin{tabular}{LL}
%   \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
%   \operator{even} (0) = \top                        &  \operator{odd}(0) = \bot\\
%   \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
% \end{tabular}

% One option is to prove that the following function is
% equivalent to the above
% \begin{align*}
%   & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
%   & \operator{even-odd} (0) = (\top, \bot)\\
%   & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
% \end{align*}
% which is clearly a catamorphism $\cata{a}$ with $a (\Nil) = (\top, \bot)$ and
% $a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in
% (\ref{eq:cata}).  Now we are left to show they are equivalent. Well, it is easy
% to do so since $\operator{even} = \pi_{1} \comp \operator{even-odd}$
% $\operator{odd} = \pi_{2} \comp \operator{even-odd}$. The proof that these two
% pairs of functions are the same is an easy induction over the natural numbers.

% This is an instance of a more general phenonema whose argument uses the concept
% of an adjunction~\cite{MacLane71}.  For this particular example, the argument is
% as follows.  There is an isomorphism between pairs of maps
% $(A \to B_{1}, A \to B_{2})$ and maps $(A \to B_{1} \times B_{2})$.  This is
% given by $\Phi (f,g) = \pair{f}{g}$ and
% $\Psi(f) = (\pi_{1} \comp f, \pi_{2} \comp f)$ where the projections $\pi$ are
% part of the counit of the adjunction between the diagonal functor
% $\Delta X = (X,X)$ and the product $\times$.  Once this correspondence is
% established we can implement the map $A \to B_{1} \times B_{2}$ as an
% hylomorphism and derive the original and more complex recursion scheme.

% % In general an adjunction is a natural correspondence of maps
% % \[
% %   LA \to B \iff A \to RB
% % \]
% % where $L$ and $R$ are called adjoints.

% However, formalising adjunctions in Coq would require to formalise a
% non-negligible amount of category theory which goes beyond the scope of the
% present work. In this paper, we will instead derive the recursion schemes as
% needed and implement them as hylos.
% % \subsubsection{Course-of-Value Recursion}
% % The knapsack algorithm is defined as following in pseudo-Haskell code:
% % \begin{align*}
% %   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
% %   & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
% %     [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
% % \end{align*}
% % %
% % with dynamic programming knapsack can be rephrased like this
% % \begin{align*}
% %   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
% %   & \operator{knapsack}\;wvs\;c= table !! c \\
% %   & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
% %   & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
% % \end{align*}

% % The recursion scheme is justified by solving a recursive equation.  First we
% % define an operation to create a memory table representing the list of capacities
% % $[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
% % for constructing the naturals $\Nat = \mu D$ the we define the generating
% % function by induction on the naturals
% % \begin{align*}
% %   & \operator{fan} : \Nat \to D_{\infty}\Nat\\
% %   & \operator{fan} (0)    = \pair{0}{\inl}\\
% %   & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
% % \end{align*}
% % The recursion scheme from a comonad is then defined as follows
% % \begin{align*}
% %   & \operator{rsfc} : \Nat \to A\\
% %   & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
% % \end{align*}
% % At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
% % Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
% % \begin{align*}
% %   & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
% %   & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
% %   & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
% %   & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
% % \end{align*}
% % and now we can define the algebra over the reals
% % \begin{align*}
% %   & \operator{alg} : DD_{\infty} \R \to \R\\
% %   & \operator{alg}\; (\inl)  = 0\\
% %   & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
% % \end{align*}

% % It remains to show how an $\operator{rsfc}$ is implemented in terms of
% % hylomorhisms.

% % \paragraph{Implementation into hylomorphisms}
% % In order to implement rsfcs as hylos we need a parametric distributive law
% % \[
% %   \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
% % \]
% % This can be defined as
% % $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
% % is instrumental to be able to lift coalgebras in the following sense. Give a
% % $G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
% % $G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
% % $G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
% % on $GX$
% % \begin{align*}
% %   & G_{\lambda}(c) : GA \to G_{\infty} GA\\
% %   & G_{\lambda}(c) = \lambda \comp G (c)
% % \end{align*}
% % where $c : C \to GC$.

% % Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
% % convert it into an algebra for the carrier $G_{\infty} A$ by
% % \begin{align*}
% %   & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
% %   & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
% % \end{align*}
% % where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
% % infinite constant list.

% % At this point the hylo is simply defined by
% % $\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
% % \begin{align*}
% %   & \operator{rsfc} : \mu G \to A \\
% %   & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
% % \end{align*}

% % % \subsection{Catamorphisms and Anamorphism}
% % % A catamorphism is the unique program defined as
% % % \[
% % %   x = a \comp \fmap{x} \comp \InOp
% % % \]
% % % Can be implemented:
% % % \[
% % %   \cata{a} = \hylo{\InOp}{a}
% % % \]

% % % An anamorphism is the unique program satisfying the equation
% % % \[
% % %   x = \OutOp \comp \fmap{x} \comp{c}
% % % \]
% % % The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
% % % corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
% % % indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.

% % \paragraph{Free Monads}
% % Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
% % solution to the following domain equation
% % \[
% %   D^{*}A = A + DD^{*}A
% % \]
% % where $\eta_{A} : A \to D^{*}A$ is the left injection and
% % $\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
% % $a : DA \to A$ there is a unique solution to the following recursive equation
% % \[
% %   x \comp \call_{A} = a \comp D(x)
% % \]
% % that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
% % the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
% % defined as $\mu = \catafree{\call}$.

% % \paragraph{Cofree Comonads}
% % Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
% % greatest solution to the following domain equation
% % \[
% %   D_{\infty}A = A + DD_{\infty}A
% % \]
% % where $\head : D_{\infty}A \to A$ is the first projection and
% % $\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
% % Given an algebra $c : A \to DA$ there is a unique solution to the following
% % recursive equation
% % \[
% %   \tail_{A} \comp x = D(x) \comp c
% % \]
% % that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
% % can extend $\tail$ to the comultiplication for the comonad
% % $\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
% % $\delta = \anacofree{\tail}$.

% % \subsection{Accumulators}
% % For an algebra $a : D(A^{P})\times P \to A$:
% % \begin{align*}
% %   & g : \mu D \times P \to B\\
% %   & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
% % \end{align*}

% % \paragraph{Hylo Implementation}
% % Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
% % solution to the following equation
% % \begin{align*}
% %   & f : \mu D \to A^{P}\\
% %   & f = a \comp \fmap (f) \comp \InOp
% % \end{align*}
% % thus $f = \hylo{\InOp}{a}$.

% % The complex recursion scheme is then
% % $g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.

% % \subsection{Mutu-Hylo}
% % For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
% % $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
% % \begin{align*}
% %   & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
% %   & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
% % \end{align*}

% % \paragraph{Hylo Implementation}
% % Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
% % \begin{align*}
% %   & f : \mu D \to A_{1} \times A_{2}\\
% %   & f = a \comp \fmap{f} \comp \InOp
% % \end{align*}
% % for which the solution is an hylo $f = \hylo{\InOp}{a}$.

% % The mutumorphism can be implemented as
% % $(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.

% % \paragraph{Coq Implementation}
% % Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
% % $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
% % We define the two recursive functions  $f_{1}$ and $f_{2}$ as

% % \[
% %   f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
% % \]
% % % A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$

% % % For example, for the functor $(1 + -)$ an algebra over this functor is a
% % % function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
% % % $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
% % % \[
% % %   D_{\infty} A \iso A \times (1 + D_{\infty})
% % % \]
% % % which is the the type of non-empty possibly infinite lists with elements in $A$

% % % First we have to assume a distributive law
% % % $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
% % % $a : D(D_{\infty} A) \to A$,
% % % \begin{align*}
% % %   & g : \mu D \to A\\
% % %   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
% % % \end{align*}
% % % where
% % % $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
% % % is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
% % % type
% % % \[
% % %   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
% % % \]
% % % which is the type of a $G$-algebra used to initialise the memory table.

% % % Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
% % % what is the unique solution of the following equation
% % % \begin{align*}
% % %   & f : \mu F \to G_{\infty} A\\
% % %   & f = a \comp \fmap{f} \comp \InOp
% % % \end{align*}
% % % The solution to this equation can be found again using an hylo
% % % $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
% % % can be recovered from this by setting $g = \head \comp f$

% % % \subsubsection{Histomorphism}
% % % The definition of $\lambda$ is:
% % % \begin{align*}
% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
% % % \end{align*}
% % % For example, for $GX = 1 + X$ we have to define
% % % \begin{align*}
% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
% % % \end{align*}
% % % % \begin{align*}
% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
% % % % \end{align*}
% % % % Defining the algebra
% % % % \begin{align*}
% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
% % % %   & \operator{alg}\; Zero           = 0\\
% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
            % % % % \end{align*}
\section{Recursion Schemes}
The structure of the a certain data is very similar to the structure of an
algorithm which processes that data. Recursion schemes aim at analyzing these
analogies.

@@

For example, in $\Set$, we can define the set of lists
$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument,
however, there is a more general way of proving these type of definitions are
well-defined.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
where $A$ is an object of the category called the \emph{carrier} of the algebra
and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$. By the property of initial $F$-algebras, for
any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
pronounced ``\emph{catamorphism}''.

>>>>

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\catafree}[1]{\Lbanana #1 \Rbanana}
\newcommand{\anacofree}[1]{\Llens #1 \Rlens}
\newcommand{\hylo}[2]{\cata{#1 \to #2}}
\newcommand{\cohylo}[2]{\ana{#1 \to #2}}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\head}{\operator{head}}
\newcommand{\inl}{\operator{inl}}
\newcommand{\inr}{\operator{inr}}
\newcommand{\tail}{\operator{tail}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Coalg}{\text{-CoAlg}}
\newcommand{\Free}{\text{Free\xspace}}
\newcommand{\fmap}[1]{\text{fmap}\;#1}

@@

\newcommand{\Str}[1]{\operator{Str}(#1)}
\newcommand{\List}[1]{\operator{List}(#1)}
\newcommand{\nil}{\operator{nil}}
\newcommand{\cons}{\operator{cons}}
\newcommand{\foldr}{\operator{foldr}}

@@

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

@@

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the work
in mechanising recursion schemes focuses on a narrow subset of the known
recursion schemes due to termination issues, or do not focus on proving
algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.
Indeed, most of the mechanisations of generic
recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
on the \emph{unfolds}. In general, Hinze et al.~\cite{HinzeWG15} has proven that
all known  recursion schemes are instances of a  divide-and-conquer algorithm
which goes under the name of \emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To
date, no work has mechanised hylomorphisms in the Coq proof assistant, together
with their algebraic laws.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration is what allows the execution of code that has been verified in
Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
al.'s approach, extraction (1) does not preserve the recursive structure of
common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
the generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

@@

\section{Recursion Schemes}
The structure of the a certain data is very similar to the structure of an
algorithm which processes that data. This relationship manifests in the form of
structured recursion schemes which are widely used in functional languages such
Haskell. Canonical examples are folds (catamorphisms), which consume data, and
unfolds (anamorphisms), which produce it. While some implementations of
recursion schemes like $\operator{foldr}$ in Haskell as specific to a particular
data structure, in this case $\operator{List}$s, we can generalise further these
ideas to account for generic (co)inductive data types and generic algorithms
operating on them. This is done by looking at recursion schemes from the point
of view of category theory.

@@

For example, in $\Set$, we can define the set of lists
$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument,
however, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
where $A$ is an object of the category called the \emph{carrier} of the algebra
and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$. By the property of initial $F$-algebras, for
any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
pronounced ``\emph{catamorphism}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{\alpha} \iff f = \alpha \circ F(f) \circ \InIso
\end{equation}
In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
catamorphism.
